---
title: "p2_g2_report"
date: "2026-02-19"
output: pdf_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(ggplot2)
library(gtsummary)
library(gridExtra)
library(ggridges)
library(maps)
library(gt)
library(tigris)
library(sf)
library(gamm4)
library(mgcv)
library(readxl)
library(lme4)
library(lmerTest)
library(glmmTMB)
library(broom.mixed)
library(performance)
library(car)
library(fixest)
library(broom.helpers)
library(purrr)

```

\newpage

# Appendix 

## EDA Section

```{r, include = FALSE}

tract <- read_csv("tract_cbsa.csv") %>% 
  mutate(dens_mult = dens*1000000)

# Long format race data
race_long <- tract %>%
  select(pct_white, pct_black, pct_asian, pct_hispanic, pm25) %>%
  pivot_longer(
    cols = starts_with("pct"),
    names_to = "race",
    values_to = "percent") %>% 
  mutate(race = recode(race,
                       pct_white = "White",
                       pct_black = "Black",
                       pct_asian = "Asian",
                       pct_hispanic = "Hispanic"))

```

**Table 1**: Census Tract Characteristics, by RUCA Categorization -- Median (Q1, Q3)

```{r}

table1 <- tract %>% 
  tbl_summary(include = c(pm25, hyads, bc, total, dens_mult, pct_pov, 
                          pct_white, pct_black, pct_asian, pct_hispanic),
              by = ruca_agg, 
              label = list(
                pm25 ~ "PM 2.5",
                hyads ~ "Coal Power Plant Impact (HyADS)",
                bc ~ "Black Carbon",
              #  area ~ "Land Area (m)",
                total ~ "Total Population",
                dens_mult ~ "Population Density (per km^2)",
                pct_pov ~ "Percent in Poverty",
                pct_white ~ "Percent White",
                pct_black ~ "Percent Black",
                pct_asian ~ "Percent Asian",
                pct_hispanic ~ "Percent Hispanic"),
             # statistic = all_continuous() ~ "{mean} ({sd})",
              digits = all_continuous() ~ 2, 
              missing = "no") %>% 
  add_p() %>% 
  add_overall() %>% 
  bold_labels() %>% 
 # modify_caption("Census Tract Characteristics, by RUCA Categorization -- Median (Q1, Q3)") %>% 
  as_kable()

table1

# table1 <- table1 %>% as_gt()
# gtsave(table1, "results/table1.png")

```


```{r, include = FALSE}
  
tract_ridges <- tract %>%
   filter(!is.na(pm25), !is.na(pct_white)) %>%
   mutate(
     white_group = ntile(pct_white, 5),
     white_group = factor(
       white_group,
       levels = 1:5,
       labels = c("0-20%", 
                  "20-40%",
                  "40-60%",
                  "60-80%", 
                  "80-100%")))

figure1 <- tract_ridges %>% ggplot(aes(x = pm25, y = white_group, fill = white_group)) +
   geom_density_ridges(
     alpha = 0.85,
     scale = 1.2,
     color = "white") +
   theme_minimal() +
   labs(
     x = expression(PM[2.5]~"Concentration (µg/m"^3*")"),
     y = "% White",
     title = "Distribution of PM2.5 by % of White Census Tract Population") +
   theme(legend.position = "none", 
         plot.title = element_text(size = 12))

# ggsave(filename = "results/figure1.png", plot = figure1, 
 #       width = 8, height = 5, dpi = 300)


tract_ridges <- tract %>%
   filter(!is.na(bc), !is.na(pct_white)) %>%
   mutate(
     white_group = ntile(pct_white, 5),
     white_group = factor(
       white_group,
       levels = 1:5,
       labels = c("0-20%", 
                  "20-40%",
                  "40-60%",
                  "60-80%", 
                  "80-100%")))

figure_bc <- tract_ridges %>% ggplot(aes(x = bc, y = white_group, fill = white_group)) +
   geom_density_ridges(
     alpha = 0.85,
     scale = 1.2,
     color = "white") +
   theme_minimal() +
   labs(
     x = "Total Black Carbon Concentration",
     y = "% White",
     title = "Distribution of Black Carbon Concentration by % of White Census Tract Population") +
   theme(legend.position = "none", 
         plot.title = element_text(size = 12))

# HyADS 

tract_ridges <- tract %>%
   filter(!is.na(hyads), !is.na(pct_white)) %>%
   mutate(
     white_group = ntile(pct_white, 5),
     white_group = factor(
       white_group,
       levels = 1:5,
       labels = c("0-20%", 
                  "20-40%",
                  "40-60%",
                  "60-80%", 
                  "80-100%")))

figure_hyads <- tract_ridges %>% ggplot(aes(x = hyads, y = white_group, fill = white_group)) +
   geom_density_ridges(
     alpha = 0.85,
     scale = 1.2,
     color = "white") +
   theme_minimal() +
   labs(
     x = "HyADS (Impact of Coal-Fired Power Plants)",
     y = "% White",
     title = "Distribution of HyADS by % of White Census Tract Population") +
   theme(legend.position = "none", 
         plot.title = element_text(size = 12))

figure_hyads

# ggsave(filename = "results/figure_hyads.png", plot = figure_hyads, 
#        width = 8, height = 5, dpi = 300)

```

**Figure 1**: Distribution of PM2.5 by Rural-Urban Commuting Area and Code

```{r, fig.align='center', fig.width=6, fig.height=4, warning = FALSE, message = FALSE}

ruca_1 <- tract %>% ggplot(aes(x = pm25, y = ruca_agg, fill = ruca_agg)) +
   geom_density_ridges(
     alpha = 0.85,
     scale = 1.2,
     color = "white") +
   theme_minimal() +
   labs(
     x = expression(PM[2.5]~"Concentration (µg/m"^3*")"),
     y = "RUCA",
     title = "Distribution of PM2.5 by Aggregated Rural-Urban Commuting Area") +
   theme(legend.position = "none", 
         plot.title = element_text(size = 12))

ruca_2 <- tract %>% ggplot(aes(x = factor(ruca), y = pm25, color = factor(ruca))) +
  geom_boxplot() +
  theme_minimal() + 
  labs(
     x = "RUCA Code",
     y = expression(PM[2.5]~"Concentration (µg/m"^3*")"),
     title = "Distribution of PM2.5 by Rural-Urban Commuting Area Code") +
   theme(legend.position = "none", 
         plot.title = element_text(size = 12))

grid.arrange(ruca_1, ruca_2, nrow = 2)

```

>

**For the below, can include either the combined figure or just the PM2.5 figure!**

**Figure 2**: Distribution of PM2.5 by % of White Census Tract Population

```{r, fig.align='center', fig.width=6, fig.height=4, message = FALSE, warning = FALSE}

figure1

```

```{r, fig.width=8, fig.height=11, fig.align='center', message = FALSE}

grid.arrange(figure1, figure_bc, figure_hyads, nrow = 3)

```


**Figure 3**: 
MAPS I will add them


**Figure 4**: Racial Composition by PM2.5 Exposure Percentile

```{r, fig.align='center', fig.width=8, fig.height=5, message = FALSE, warning = FALSE}

pm_long <- race_long %>%
  filter(!is.na(pm25)) %>% 
  mutate(pm25_percentile = ntile(pm25, 5)) %>% 
  mutate(pm25_percentile = factor(
    pm25_percentile,
       levels = 1:5,
       labels = c("0-20%", 
                  "20-40%",
                  "40-60%",
                  "60-80%", 
                  "80-100%")))

figure_pm_bar <- ggplot(pm_long, aes(x = factor(pm25_percentile),
                 y = percent,
                 fill = race)) +
  geom_bar(stat = "identity", position = "stack") +
  # facet_wrap(~ ruca_agg) +
  labs(x = "PM2.5 Percentile",
       y = "Proportion",
       fill = "Race",
       title = "Racial Composition by PM2.5 Exposure Percentile") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12))

figure_pm_bar

```


**Figure 5**: PM2.5 vs. of % Race of Census Tract Population

```{r, fig.align='center', fig.width=8, fig.height=5, message = FALSE, warning = FALSE}

figure3 <- race_long %>% 
  ggplot(aes(x = percent, y = pm25, color = race)) +
  geom_point(alpha = 0.3) +
  geom_smooth(color = "black") +
  facet_wrap(~race) +
  theme_minimal() +
  labs(x = "% of Census Tract Population",
       y = PM[2.5]~"Concentration (µg/m"^3*")") +
  theme(plot.title = element_text(size = 12))

figure3

```

# Modeling
## Approach 1: Nested LMM

## Approach 2: Cross classified Multilevel modeling (CCMM)
```{r}
############################# Data assessment ###########################

cbsa_data  <- tract %>% mutate(state_fips = factor(state_fips),
state_abb = factor(state_abb) ,
county_fips = factor(county_fips),
geoid = factor(geoid)    ,
cbsa_id = factor(cbsa_id)   ,
#ruca = factor(ruca)     ,
ruca_agg = factor(ruca_agg),
density_log = log(dens_mult))


## Check the distinct count of states, cbsa, census tracts and county

n_distinct(cbsa_data$state_abb) ## No Massachusetts, Wyoming
n_distinct(cbsa_data$cbsa_id)
n_distinct(cbsa_data$county_fips)
n_distinct(cbsa_data$geoid)

## Check for missingness in outcome

sum(is.na(cbsa_data$pm25)) # 1 missing

## Check the distribution of outcome

hist(cbsa_data$pm25) ## normal data


## Check whether data is actually nested or cross classified

### 1. Does each tract belong to one CBSA?

cbsa_data %>%
  group_by(geoid) %>%
  summarise(n_cbsa = n_distinct(cbsa_id)) %>%
  filter(n_cbsa > 1) ## Truly nested

### 2. Does each CBSA belong to only ONE state
cbsa_data %>%
  group_by(cbsa_id) %>%
  summarise(n_states = n_distinct(state_fips)) %>%
  filter(n_states > 1) ## 43 cbsa_ids appear in more than 1 state; not truly nested


### Plot to show non-nested CBSAs
cbsa_data %>%
  group_by(cbsa_id) %>%
  summarise(n_states = factor(n_distinct(state_fips))) %>% ggplot(aes(n_states)) + geom_bar() + labs(
    #title = "Count of CBSA ",
    x = "Number of states",        # x-axis label
    y = "CBSA count"    # y-axis label
  ) + theme_minimal() + scale_fill_viridis_d()



  labs(
    title = "Example Bar Plot",
    x = "Category",        # x-axis label
    y = "Count / Value"    # y-axis label
  ) +
  theme_minimal()


## Check correlations and multicolinearity

### Correlation
#cor(cbsa_data[, c("pct_white","pct_hispanic","pct_black", "pct_pov", "dens_mult", "bc", "pm25", "ruca", "hyads")], use = "complete.obs")


### Multicolinearity using non-nested model

#vif(lm(pm25 ~ pct_white + pct_hispanic + pct_black + pct_pov + ruca_agg + dens + bc + hyads, data = cbsa_data))

```


```{r fig.height=15, fig.width=10,  fig.align='left'}

# Justification:

# Each tract belongs to one CBSA
# Each tract belongs to one state
# But CBSAs and states are not truly nested within each other i.e., cbsa crosses state boundaries
      ## So tracts are simultaneously classified by:
            ## Tract ∈ CBSA
            ## Tract ∈ State
# Can compute between cbsa and state variances as well as within tract variances

# Limitations:

## Assumes normal random effects
## It doesn't handle spatial autocorrelation
## If state doesn't contribute any variance it becomes an unneccessarily model complex with no benefit


## Nested LMM 
cc_model1 <- lmer(
  pm25 ~ pct_white*pct_pov  + ruca_agg + density_log + 
    #(1 | cbsa_id) +
    (1 | state_abb/cbsa_id),
  data = cbsa_data
)


## CCMM
cc_model <- lmer(
  pm25 ~ pct_white*pct_pov +  ruca_agg + density_log + 
    (1 | cbsa_id) +
    (1 | state_abb),
  data = cbsa_data)

#summary(cc_model)

 #tidy(cc_model,  conf.int = TRUE) %>% knitr::kable() 

## Compare models
 anova(cc_model1, cc_model)

######################## Model diagnostics ##############################

 library(performance)

compare_performance(cc_model1, cc_model)

# save model diagnostics plot
png("cclmm_diagnostics.png", width = 14*300, height = 14*300, res = 300)  
check_model(cc_model)

dev.off()

## check distribution of random effects

ranef_vals <- ranef(cc_model)
hist(ranef_vals$cbsa_id[[1]], main="", xlab="Random effect: CBSA")
hist(ranef_vals$state_abb[[1]], main="", xlab="Random effect: State")

##  Model metrics
model_performance(cc_model)


################### create and save report-ready results table ##########

model_tbl <- tbl_regression(
  cc_model,
  intercept = TRUE,
  conf.int = TRUE,
  estimate_fun = function(x) sprintf("%.4f", x),
  label = list(pct_white ~ "Percent White in pop",
               pct_pov ~ "Percent pop in pov",
               ruca_agg ~ "Rural-Urban Commuting Area",
               density_log ~ "Log density (people/sq km)")
) %>%
  bold_labels() %>%  modify_header(label ~ "Term") %>%  # rename 'Characteristic' -> 'Term'
  modify_caption("**Cross-Classified Mixed Effects Model for PM2.5**") 


model_tbl

 as_gt(model_tbl) %>%
  gt::gtsave("model_ccmm_results.png")


```

## Approach 3: Multiway cluster adjusted SE (CRSE)
```{r}

# Justifications:
## This computes two-way clustered standard errors:
## Clustered by CBSA
## Clustered by State
## Correcting for overlapping membership

# Limitations
## Does not estimate variance at CBSA or state level.
  ## Cannot ICC: cannot attribute any variation to CBSA or state
## Does NOT add CBSA or state intercepts
## No cluster-level estimates
## Does NOT change coefficient estimates
## Leaves the model as simple OLS.
  ## Only adjusts standard errors
  ## Only adjusts the variance-covariance matrix.
  ## Assumes independence across clusters

cr_model <- feols(
  pm25 ~ pct_white*pct_pov + ruca_agg + density_log,
  data = cbsa_data,
  cluster = ~ cbsa_id + state_abb)

summary(cr_model)

########################### Model diagnostics ##########################

### compare cluster SEs

library(data.table)

coefs <- names(coef(cr_model))

se_table <- data.table(
  term = coefs,
  cluster_cbsa   = sqrt(diag(vcov(cr_model, cluster = ~cbsa_id))),
  cluster_state = sqrt(diag(vcov(cr_model, cluster = ~state_abb))),
  cluster_both = sqrt(diag(vcov(cr_model, cluster = ~cbsa_id + state_abb)))
)

se_table <-  se_table %>% mutate(term = case_when(term == "pct_white" ~ "Percent White in pop",
               term == "pct_pov" ~ "Percent pop in pov",
               term == "ruca_agg" ~ "Rural-Urban Commuting Area",
               term == "density_log" ~ "Log density (people/sq km)",
               term == "pct_white:pct_pov" ~ "White X poverty", TRUE ~ term),
               cluster_cbsa = round(cluster_cbsa, 4),
               cluster_state = round(cluster_state, 4),
               cluster_both = round(cluster_both, 4))

se_table

gt_table <- gt(se_table)

# Save as PNG
gtsave(gt_table, "se_table.png")

## Quick model comparison using AIC and BIC

AIC(cr_model)
BIC(cr_model)
AIC(cc_model)
BIC(cc_model)

################### create and save report-ready results table ##########

cr_tbl <- tbl_regression(
  cr_model,
  intercept= TRUE,
  conf.int = TRUE,
  estimate_fun = function(x) sprintf("%.4f", x),
  label = list(pct_white ~ "Percent White in pop",
               pct_pov ~ "Percent pop in pov",
               ruca_agg ~ "Rural-Urban Commuting Area",
               density_log ~ "Log density (people/sq km)"
               )
) %>%
  bold_labels() %>%  modify_header(label ~ "Term") %>%  # rename 'Characteristic' -> 'Term'
  modify_caption("**Multiway cluster adjusted SE Model for PM2.5**") 

as_gt(cr_tbl) %>%
  gt::gtsave("model_crse_results.png")


```

## Approach 4: Generalized additive mixed models (GAMMs)
```{r}

# In GAMs beta coefficients from Linear Regression are replaced with a flexible function (splines) that allows nonlinear relationships 

# Splines allow us to model non-linear relationships for each feature. The sum of many splines forms a GAM.The result is a highly flexible model which still has some of the explainability of a linear regression.

## GAMM Using nlme for multilevel modeling part
cbsa_data$state_cbsa <- interaction(cbsa_data$state_abb, cbsa_data$cbsa_id)

gam_nlm <- gamm(pm25~s(pct_white) + pct_white:pct_pov + pct_pov + ruca_agg + density_log, data= cbsa_data,random = list(state_abb = ~1,
                state_cbsa = ~1), method = "REML")

########################### Model diagnostics ##########################

## Use these GAMMs to obtain comparison between linear and smooth models (These GAMMS LMM part based on lme4)
smooth_model <- gamm4(pm25~s(pct_white) + pct_white:pct_pov + pct_pov + ruca_agg + density_log, data= cbsa_data,random =~(1|state_abb / cbsa_id))

linear_model <- gamm4(pm25~pct_white + pct_white:pct_pov + pct_pov + ruca_agg + density_log, data= cbsa_data,random =~(1|state_abb / cbsa_id))


anova(linear_model$mer, smooth_model$mer, test="Chisq")

# Plot smooth terms to assess non-linearity of relationship between pm2.5 and pct_white
png("gam_smooth_plot.png", width = 800, height = 600, res = 120)

plot(gam_nlm$gam, shade = TRUE, pages = 1)

# Close device
dev.off()


logLik(gam_nlm$lme)
AIC(gam_nlm$lme)
BIC(gam_nlm$lme)
gam_nlm$lme$sigma


################### create and save report-ready results table ##########
 ### Fixed effect part

gam_tbl <- tbl_regression(
  gam_nlm$lme,
  intercept= TRUE,
  conf.int = TRUE,
  estimate_fun = function(x) sprintf("%.5f", x),
  label = list(`Xs(pct_white)Fx1` ~ "Percent White in pop",
               Xpct_pov ~ "Percent pop in pov",
               Xruca_aggsuburban ~ "Urbanicity - Suburban",
               Xruca_aggurban ~ "Urbanicity - Urban",
               `Xdensity_log` ~ "Log density (people/sq km)",
               `X(Intercept)` ~ "Intercept",
               `Xpct_pov:pct_white` ~ "pct White X pct pov"
               #bc ~ "Black carbon conc",
               #hyads ~ "HyADS"
               #pct_white:pct_pov ~ "White X pov",
               #bc:hyads ~ " Black carbon X HyADS"
               )
) %>%
  bold_labels() %>%  modify_header(label ~ "Term") %>%  # rename 'Characteristic' -> 'Term'
  modify_caption("**Non-linear GAMM for PM2.5**") 

as_gt(gam_tbl) %>%
  gt::gtsave("model_gamm_results.png")

## smooth terms part

smooth_df <- as.data.frame(summary(b$gam)$s.table)
smooth_df$term <- rownames(smooth_df)

smooth_df <- smooth_df[, c("term", "edf", "F", "p-value")]
colnames(smooth_df) <- c("term", "edf", "statistic", "p.value")

smooth_df <-  smooth_df %>% mutate(p.value = case_when(p.value==0~"<0.001", TRUE~as.character(p.value)))


smooth_df %>%
  gt() %>%
  tab_header(title = "Smooth Terms") %>%
  gtsave("smooth_terms_gt.png")


```

